---
title: "Marginal standardization"
output:
  html_document: default
date: "2024-10-28"
---

```{r load-libraries}

#uncomment and compile the install.packages() code if you need to install any of
#these libraries.
# install.packages("tidyverse")
# install.packages("broom")
# install.packages("readxl")
# install.packages("magrittr")

library(tidyverse)
library(broom)
library(readxl)
library(magrittr)
```

```{r import-data}
photo_dat <- read_xls("../data/phototherapy.xls")
```

```{r look-at-data}
names(photo_dat)
str(photo_dat)
head(photo_dat)
```

**Description of dataset**

* These data were downloaded from the textbook course site 
https://regression.ucsf.edu/second-edition/data-examples-and-problems

* From UCSF authors!

* These data look for the effect of phototherapy on neonatal jaundice within 
babies born in the Northern California Kaiser Permanente system over a 10-year
period, 1995-2004.

**Exposure of interest**

```{r}
table(photo_dat$phototherapy)

##22% (~1 in 5) infants exposed to phototherapy
round(prop.table(table(photo_dat$phototherapy))*100, 2)
```

**Outcome of interest**

```{r}
table(photo_dat$over_thresh)

# 0.62 (or 62 per 10,000 infants) have the outcome of interest 
# --> rare event
round(prop.table(table(photo_dat$over_thresh))*100, 2)
```

**Crude risk among infants with phototherapy**

```{r}
photo_dat %>% 
  filter(phototherapy == "yes") %>% 
  count(over_thresh) %>% 
  mutate(risk = round((n/sum(n))*100, 2))
```

```{r}
photo_dat %>% 
  filter(phototherapy == "no") %>% 
  count(over_thresh) %>%
  mutate(risk = round((n/sum(n))*100, 2))
```

Summary: 70 out of every 10,000 infants develop the outcome among those without phototherapy compared to 33 out of every 10000 infants with photo-therapy.

*IF* (and that is a BIG IF) the crude association could be interpreted causally,
then we could say that phototherapy averted 70-33 = 37 cases of the outcome for
every 10,000 infants treated with phototherapy.

Remember this is a crude (unadjusted) effect estimate so there are **strong** assumptions required to interpret this causally, such as the treated and 
untreated populations being exchangeable. 

```{r data-cleaning}
photo_dat <- photo_dat %>% mutate(over_thresh2 = if_else(over_thresh == "no", 0, 1))
photo_dat <- photo_dat %>% mutate(phototherapy2 = if_else(phototherapy == "no", 0, 1))

table(photo_dat$qual_TSB)
#categorical variable with 6 levels
str(photo_dat$qual_TSB)
#this variable is stored as a string variable type 

#####################################
######## tidy TSB variable ##########
#####################################

#to use this variable in R, we need to recode it as a factor variable and can do this in two steps:
photo_dat <- photo_dat %>% 
  mutate(qual_TSB2 = as.factor(qual_TSB), #store as a factor variable
         #reorder the factor levels so that "0 to <1" is the referent category, followed by the other levels:
         qual_TSB_f = fct_relevel(qual_TSB2, 
                                  "0 to <1", 
                                  "-3 to <-2",
                                  "-2 to <-1",
                                  "-1 to <0", 
                                  "1 to <2", 
                                  "2 to <3"))

photo_dat %<>% select(-qual_TSB2) #remove variable used only for coding

########################################
######## tidy gestational age ##########
########################################

#to use this variable in R, we need to recode it as a factor variable and can do this in two steps:
photo_dat <- photo_dat %>% 
  mutate(gest_age2 = as.factor(gest_age), #store as a factor variable
         #reorder the factor levels so that 40 weeks is the referent category, followed by the other levels:
         gest_age_f = fct_relevel(gest_age2, 
                                  "40", "35", "36", "37", "38", "39", ">=41"))

photo_dat %<>% select(-gest_age2) #remove variable used only for coding

########################################
######## tidy current age ##########
########################################

#to use this variable in R, we need to recode it as a factor variable and can do this in two steps:
photo_dat <- photo_dat %>% 
  mutate(age_days2 = as.factor(age_days), #store as a factor variable
         #reorder the factor levels so that 40 weeks is the referent category, followed by the other levels:
         age_days_f = fct_relevel(age_days2,
                                  "48 to <72h", "<24 h", "24 to <48h", 
                                  "72 to <96h", ">= 96h"))

photo_dat %<>% select(-age_days2) #remove variable used only for coding

photo_dat %<>% mutate(male_f = factor(male))
```

Before we run an adjusted model, below, suppose that we made a DAG and 
identified the following variables as hypothesized confounders of the 
relationship between the exposre of interest (`phototherapy2`) and the outcome
of interest (`over_thresh2`): qual_TSB_f, gest_age_f, birth_wt, age_days_f, 
male_f.

```{r adjusted-model}
adjusted_model <- glm(over_thresh2 ~ phototherapy2 + qual_TSB_f + gest_age_f + 
                          birth_wt + age_days_f + male_f, data = photo_dat, 
                      family = "binomial")

summary(adjusted_model)
tidy(adjusted_model)
```

```{r set-exposure-for-parallel-worlds}
photo_dat %<>% mutate(pred_obs  = predict(adjusted_model, type = "response"))

# Create  a duplicated version of these data, except **SETTING** everyone's exposure
# value to 0
photo_dat_0 <- photo_dat %>% mutate(phototherapy2 = 0)

# Create another duplicated version of these data, except **SETTING** everyone's exposure
# value to 1
photo_dat_1 <- photo_dat %>% mutate(phototherapy2 = 1)

#take a look at the first six rows of each dataset and ensure exposure is equal 
#to zero as expected:
head(photo_dat_0)

#take a look at the first six rows of each dataset and ensure exposure is equal 
#to one as expected:
head(photo_dat_1)
```

```{r predict-for-each-world}
photo_dat %<>% 
  mutate(pred_set_untreated = predict(adjusted_model, type = "response",
                                      newdata = photo_dat_0))

photo_dat %<>% 
  mutate(pred_set_treated = predict(adjusted_model, type = "response",
                                    newdata = photo_dat_1))

photo_dat %>% head() %>% View()
```

```{r}
# given that the predicted probabilities are so small, multiple them by 1000:
photo_dat %<>% mutate(pred_set_untreated2 = pred_set_untreated*10000,
                      pred_set_treated2 = pred_set_treated*10000)


# calculate the average (aka marginal) risk ratio (rr), odds ratio (or) and risk 
# difference (rd)
photo_dat %>% summarise(risk_set_untreated = mean(pred_set_untreated),
                        risk_set_treated = mean(pred_set_treated),
                        odds_treated = risk_set_treated/(1-risk_set_treated),
                        odds_untreated = risk_set_untreated/(1-risk_set_untreated),
                        rd = risk_set_treated - risk_set_untreated,
                        rd_per1k = rd * 1000,
                        rr = risk_set_treated/risk_set_untreated,
                        or = odds_treated/odds_untreated) %>% View()
```
